---
title: "Optional Dependencies"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

There are a variety of packages that `uwot` will make use of if you install them
(and load them), but that you don't *need* so are optional. These include:

* [RcppHNSW](https://cran.r-project.org/package=RcppHNSW) -- used for nearest 
neighbor search. Once installed and loaded, you can specify `nn_method = "hnsw"`
in `uwot::umap` as long as your `metric` is either `"euclidean"`, `"cosine"` or
`"correlation"`. This should be a bit faster than the default of Annoy in most
cases. If you use `uwot::umap2` then you will get HNSW by default without having
to specify `nn_method`.
* [rnndescent](https://cran.r-project.org/package=rnndescent)  -- used for 
nearest neighbor search. Once installed and loaded, you can specify `nn_method =
"nndescent"` in `uwot::umap`. `rnndescent` can handle many metrics, so see its
[documentation](https://jlmelville.github.io/rnndescent/articles/metrics.html)
for more information. If you use `uwot::umap2` and do not load `RcppHNSW`, then
you will use this method by default without having to specify `nn_method`.
You can also use sparse matrices as input to `uwot::umap2`. See the
[sparse data article](https://jlmelville.github.io/uwot/articles/sparse-data-example.html)
for more details.

There is one other dependency,
[RSpectra](https://cran.r-project.org/package=RSpectra) -- used for the default
spectral initialization. If not installed, then the
[irlba](https://cran.r-project.org/package=irlba) package is used instead. In
most cases `irlba` does a fine job, but it's not as fast as `RSpectra` for
spectral initialization because `irlba` isn't designed for quite the same use
case as `RSpectra`. This has been both a core dependency and an optional
dependency at different times in the past. In general, I would like it to be an
optional dependency, but currently, an interaction with the most recent release
of `irlba` and an ABI change to the `Matrix` package is causing failures. By
making `RSpectra` a required dependency, I can ensure tests, examples and
vignettes build which unblocks various downstream packages.

My recommendation would be to install all of these (or at least `RcppHNSW`):

```{r, install dependencies}
install.packages(c("RcppHNSW", "rnndescent"))
```

and then have them (and `RSpectra`) loaded whenever you are using `uwot`.

```{r, load dependencies}
library(RSpectra)
library(RcppHNSW)
library(rnndescent)
library(uwot)
```

The following UMAP run will then use `RcppHNSW` and `RSpectra` (versus
`RcppAnnoy` and `irlba`) without you having to specify anything:

```{r, UMAP using optional dependencies}
iris_umap <- umap2(iris, n_neighbors = 30)
```
